---
title: "Tutorial 1: Objectives of Financial Reporting, Properties of Earnings and Information Asymmetry"
author: Ian D. Gow
date: 2024-09-03
date-format: "D MMMM YYYY"
format: 
  html: default
  pdf: 
    colorlinks: true
    geometry:
      - left=2.5cm
      - right=2.5cm
    papersize: a4
    mainfont: TeX Gyre Pagella
    mathfont: TeX Gyre Pagella Math
---


The tutorial contains 6 questions all of which you should prepare for.

Q1 involves a discussion of the primary objective of the financial economic business system: to pool and facilitate the allocation of resources/capital to the most productive/profitable use. The question will then begin to provide some insights into benchmarks/base rates for determining high and low profitability. This includes obtaining an understanding of (some) the properties of earnings and the use of  simple data analytics such as a histogram to visualize  and describe the distribution of business variables.      Q2 and Q3 introduces the concepts of information asymmetry to illustrate how this concept can explain a significant amount of economic resource allocation problems and associated outcome/solutions.  Q4 introduces a primary feature of financial reports that can reduce information asymmetry: accruals. Q5 and Q6 begin to introduce possible biases associated with accruals which we will be discussing in subsequent weeks. 

## Question 1. 

Below is the histogram and descriptive statistics for the distribution of ROE and Annual Share Returns for Australian ASX listed firms over the period from 2000 to 2023. 
Discuss the following.

 a) What do the descriptive statistics and histogram imply about the central tendency and statistical shape of the distribution of ROE and returns?
 b)	What are the underlying reasons for the shape of the ROE and Return distributions and why do the distributions differ?
 c)	What are the implications of the nature of the ROE and Return distributions for the use of accounting information to analyse economic performance?

For self-learning you can reproduce  these descriptive statistics and a histogram (and also a [density plot](http://www.sthda.com/english/wiki/ggplot2-density-plot-quick-start-guide-r-software-and-data-visualization)) for both the ROE and return variables for the population of both all firms and firms grouped by size. 
The dataset `ROEdata2024.csv` contains the annual return on equity (`roe`) and share returns (`return`) for Australian listed firms over the period from 2010 to 2023.
The firms have been classified into 4 groups based on size as measured by market capitalization from small to large. 
See the variable `size_group` where the smallest stocks are classified as "Micro" and the largest stocks are classified as "Large".  

```{r}
#| warning: false
library(tidyverse)
library(modelsummary)
library(tinytable)
```

```{r}
ROEData2024 <- read_csv("data/ROEData2024.csv", col_types = "cddd")
```

We start by creating a subsample that removes observations that do not meet criteria. 
Getting percentile rank by size---based on total assets `at`) and forming size groups.

```{r}
recode_size <- function(x) {
  levels <- c("Micro" = "1", "Small" = "2", "Medium" = "3",  "Large" = "4")
  fct_recode(factor(x), levels = !!!levels)
}

Benchdata <- 
  ROEData2024 |>
   filter(fyear > 1999, 
          roe>-3 & roe <3, 
          return > -3 & return < 3) |>
  group_by(fyear) |>
  mutate(size_group = recode_size(ntile(at, 4)),
         size_decile = ntile(at, 10),
         large = size_decile > 8) |> 
  ungroup()

fmt_percent <- function(x) sprintf("%.0f", x * 100)
```

Producing descriptive stats

```{r}
P10 <- function(x) quantile(x, probs = 0.10, na.rm = TRUE)
P90 <- function(x) quantile(x, probs = 0.90, na.rm = TRUE)

add_header <- function(x, cols) {
  group_tt(x, j = list("Percentiles" = cols)) |>
  style_tt(tabularray_inner = "row{1}={halign=c,},")
}
```

```{r}
#| label: tbl-ret-all
#| tbl-cap: Descriptive statistics -- ROE and returns (%)
datasummary((roe + return) 
            ~ Mean + Max + P10 + P25 + P50 + P75 + P90 + Min + SD,
            data = Benchdata, fmt = fmt_percent, digits = 0) |>
  add_header(cols = 4:8)
```

```{r}
#| label: tbl-ret-group
#| tbl-cap: Descriptive statistics -- ROE and returns by size group (%)
datasummary((roe + return) * (size_group + 1) 
            ~ Mean + Median + Min + P10 + P25 + P50 + P75 + P90 + Max + SD,
            data = Benchdata, fmt = fmt_percent) |>
  add_header(cols = 6:10)
```


## Produce histogram and density plots

First define the specific values to be shown on the $x$-axis.

```{r}
breaks <- seq(-3, 4, 0.50)
my_vector <- c(-1, -0.5, -0.25, 0, 0.25, 0.5, 1)
```

```{r}
#| label: fig-hist-ret
#| fig-cap: Histogram of Annual Returns
Benchdata |>
  ggplot(aes(return)) +
  geom_histogram(binwidth = 0.02) +
  scale_x_continuous(breaks = breaks) +
  xlab("Annual Return") + ylab("Frequency")
```

```{r}
#| label: fig-hist-roe
#| fig-cap: Histogram of ROE
Benchdata |>
  ggplot(aes(x = roe)) +
  geom_histogram(binwidth = 0.02) +
  scale_x_continuous(breaks = breaks) +
  xlab("ROE") + ylab("Frequency")
```

```{r}
#| label: fig-density-roe
#| fig-cap: Density of ROE
Benchdata |>
  ggplot(aes(x = roe)) +
  geom_density() +
  scale_x_continuous(breaks = breaks) +
  xlab("ROE") + ylab("Density")
```

```{r}
#| label: fig-density-ret
#| fig-cap: Density of returns
Benchdata |>
  ggplot(aes(x = return)) +
  geom_density() +
  scale_x_continuous(breaks = breaks) +
  xlab("Annual Return") + ylab("Density")
```


## Question 2 

The bid–ask spread (also bid–offer) is the percentage difference between the prices quoted for an immediate sale (ask/offer) and an immediate purchase (bid) for shares. 
The graph below shows over the period from 2019 to 2021 the five-day moving average of the daily percentage bid-ask spread for two Australian listed companies: Woolworths and Metcash.  

The average bid-ask spread for Woolworths is 0.55% and for Metcash is 1.72%

Using some of the background information to Woolworths and Metcash provided below explain why these bid-ask spreads are different?
Are there any economic consequences from this difference?

Woolworths is one of Australia's largest supermarket chains. 
Metcash is a wholesale distributor and supplier to a large number of independently (family owned) retailers such as the IGA stores.
Metcash thus has arguably a more complicated business model.  

|                                          |                Metcash               |   Woolworths  |
|:-----------------------------------------|:------------------------------------:|:-------------:|
| Principal operations                     | Distributor to independent retailers |  Supermarket  |
| Market capitalization                    |             $4 billion               | $43 billion  |
| Analyst coverage                         |                  13                  |       18      |
| Daily trading volume (2022)              |                 $13m                 |     $68m      |
| Dispersion in analyst revenue forecasts  |                25.45%                |      8.4%     |

## Question 3

Firms are increasingly delisting and being bought out by private equity firms. 
Use the concepts of information asymmetry to explain this.
Some PE firms in Australia include EQT, TPG Capital and BGH Capital.
A recent private equity transaction in Australia is the purchase of Crown Casino by New York based private equity firm Blackstone. 
See attached article. 

# Question 4

Telstra is Australia’s largest telecommunications company. 
Obtain [Telstra’s 2023 Annual Report](https://www.telstra.com.au/content/dam/tcom/about-us/investors/pdf-g/TEL-AR-2022-Spreads-FINAL.pdf).
Compute the level of accruals for Telstra and answer the following questions:

 - Will accruals increase or decrease information asymmetry? 
 - What type of information asymmetry will be affected? 
 - What is usually the largest accrual transaction? 
 - What do you think is the typical magnitude of accruals for a typical company? 
 - Why does the magnitude of accruals vary significantly across companies?

# Question 5

Review Woolworth's 2023 financial reports.
Find an example of the use of this report to reduce moral hazard problems. 
What is implication of this for the statistical properties of the quality of information in financial reports: bias and precision? 

# Question 6 

Find an example of a conservative financial accounting policy in the 2023 financial reports of Telstra.
In lecture we discussed that lenders may demand conservative financial reports. 
Would the shareholders of Telstra demand conservative financial reports? 
Will this type of accounting increase or decrease information asymmetry problems?





